---
title: "R4DS EDA Exercises"
format: html
execute:
  warning: false
  message: false
---

```{r}
library(tidyverse)
theme_set(theme_minimal(base_size = 12))
```

## Exercise 1

```{r}
diamonds |>
  summarise(
    total_diamonds = n(),
    diamonds_between_1000_and_5000 = sum(price >= 1000 & price <= 5000),
    diamonds_removed_by_xlim = total_diamonds - diamonds_between_1000_and_5000
  )
```

```{r}
price_histogram <- diamonds |>
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 500, boundary = 0, fill = "#2f6f73", color = "white") +
  labs(x = "Price", y = "Count")

price_histogram +
  coord_cartesian(xlim = c(1000, 5000)) +
  labs(title = "coord_cartesian()")
```

```{r}
price_histogram +
  xlim(1000, 5000) +
  labs(title = "xlim()")
```

`coord_cartesian()` zooms after plotting. `xlim()` removes data before plotting.

```{r}
diamonds |>
  ggplot(aes(x = price)) +
  geom_histogram(fill = "#4c5b8a", color = "white") +
  coord_cartesian(xlim = c(1000, 5000)) +
  labs(title = "Default bins", x = "Price", y = "Count")
```

Without `binwidth`, ggplot chooses the bins automatically.

```{r}
price_histogram +
  coord_cartesian(xlim = c(750, 2250)) +
  labs(title = "Partial bins")
```

`coord_cartesian()` can show partial bins because it crops the plot.

## Exercise 2

```{r}
diamonds |>
  summarise(
    min_price = min(price),
    q1 = quantile(price, 0.25),
    median = median(price),
    mean = mean(price),
    q3 = quantile(price, 0.75),
    max_price = max(price)
  )
```

```{r}
diamonds |>
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 500, fill = "#4c5b8a", color = "white") +
  labs(title = "Price distribution", x = "Price", y = "Count")
```

```{r}
diamonds |>
  filter(price <= 3000) |>
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 25, fill = "#8a5a44", color = "white") +
  labs(title = "Prices below 3000", x = "Price", y = "Count")
```

```{r}
diamonds |>
  distinct(price) |>
  arrange(price) |>
  mutate(
    previous_price = lag(price),
    gap = price - previous_price
  ) |>
  arrange(desc(gap)) |>
  slice_head(n = 5)
```

Prices are right-skewed. The biggest gap is between 1454 and 1546.

## Exercise 3

```{r}
diamonds |>
  filter(carat %in% c(0.99, 1.00)) |>
  count(carat, name = "number_of_diamonds")
```

```{r}
diamonds |>
  filter(carat >= 0.95, carat <= 1.05) |>
  count(carat, name = "number_of_diamonds") |>
  ggplot(aes(x = factor(carat), y = number_of_diamonds)) +
  geom_col(fill = "#3f7d4e") +
  labs(title = "Counts near 1 carat", x = "Carat", y = "Count")
```

There are many more 1.00 carat diamonds than 0.99 carat diamonds. The 1 carat threshold matters commercially.
