---
title: "Weight, Cylinders, and Fuel Economy: A Data Story from mtcars"
author: "Dario Bonfiglio"
format:
  html:
    toc: true
    theme: cosmo
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

## Objective

The `mtcars` dataset contains 32 car models and several technical variables, including fuel economy, weight, cylinders, horsepower, and transmission type. This story answers one main question:

**How much do vehicle weight and engine configuration affect fuel economy?**

To answer it, the document uses three plot types:

- a scatter plot to show the relationship between weight and fuel economy;
- a boxplot to compare fuel economy distributions by cylinder group;
- a bar chart to compare average horsepower across groups.

## Data Preparation

```{r}
library(ggplot2)
library(dplyr)
library(tibble)

ink <- "#333333"
muted <- "#555555"
paper <- "white"
grid_line <- "#E2E2DC"

cyl_palette <- c(
  "4 cylinders" = "#2A9D8F",
  "6 cylinders" = "#E9C46A",
  "8 cylinders" = "#E76F51"
)

theme_story <- function(base_size = 13) {
  theme_light(base_size = base_size) +
    theme(
      plot.background = element_rect(fill = paper, color = NA),
      panel.background = element_rect(fill = paper, color = NA),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = grid_line, linewidth = 0.4),
      plot.title = element_text(
        size = 20,
        face = "bold",
        hjust = 0.5,
        color = ink,
        margin = margin(b = 4)
      ),
      plot.subtitle = element_text(
        size = 12,
        color = muted,
        hjust = 0.5,
        margin = margin(b = 10)
      ),
      plot.caption = element_text(
        size = 9,
        color = "#666666",
        hjust = 1,
        margin = margin(t = 10)
      ),
      axis.title = element_text(size = 12, face = "bold", color = ink),
      axis.text = element_text(size = 10, color = ink),
      legend.position = "top",
      legend.title = element_text(size = 10, face = "bold"),
      legend.text = element_text(size = 9),
      legend.background = element_blank(),
      legend.key = element_blank(),
      plot.margin = margin(12, 14, 12, 14)
    )
}

theme_set(theme_story())

cars_data <- mtcars |>
  rownames_to_column("model") |>
  mutate(
    cylinders = factor(
      cyl,
      levels = c(4, 6, 8),
      labels = c("4 cylinders", "6 cylinders", "8 cylinders")
    ),
    transmission = factor(
      am,
      levels = c(0, 1),
      labels = c("Automatic", "Manual")
    ),
    weight_1000_lb = wt,
    efficiency_mpg = mpg,
    horsepower = hp
  )

engine_summary <- cars_data |>
  group_by(cylinders) |>
  summarise(
    models = n(),
    avg_mpg = mean(efficiency_mpg),
    avg_weight_1000_lb = mean(weight_1000_lb),
    avg_hp = mean(horsepower),
    .groups = "drop"
  )
```

This table sets the baseline: models are grouped by cylinder count, so we can compare average fuel economy, weight, and horsepower.

```{r}
engine_summary |>
  transmute(
    `Cylinder group` = cylinders,
    `Models` = models,
    `Average mpg` = round(avg_mpg, 1),
    `Average weight (1000 lbs)` = round(avg_weight_1000_lb, 2),
    `Average horsepower (hp)` = round(avg_hp, 0)
  ) |>
  knitr::kable(caption = "Summary statistics by cylinder group.")
```

## 1. Weight and Fuel Economy

The first chart shows the strongest relationship: lighter cars tend to get better fuel economy. Four-cylinder models are concentrated in the lighter and more efficient area of the plot.

```{r}
#| label: fig-weight-efficiency
#| fig-width: 7.2
#| fig-height: 5
#| out-width: "100%"
#| fig-cap: "Relationship between vehicle weight and fuel economy. Weight is measured in thousands of pounds; fuel economy is measured in miles per gallon."

ggplot(cars_data, aes(x = wt, y = mpg)) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    color = "#2F4858",
    fill = "#9FC5D1",
    linewidth = 1,
    alpha = 0.25
  ) +
  geom_point(
    aes(fill = cylinders, shape = transmission),
    size = 4,
    stroke = 0.9,
    color = "white",
    alpha = 0.95
  ) +
  scale_fill_manual(values = cyl_palette) +
  scale_shape_manual(values = c("Automatic" = 21, "Manual" = 24)) +
  scale_x_continuous(breaks = seq(1.5, 5.5, 0.5)) +
  scale_y_continuous(breaks = seq(10, 35, 5)) +
  coord_cartesian(xlim = c(1.4, 5.6), ylim = c(9.5, 35.5)) +
  labs(
    title = "Fuel Efficiency vs Car Weight",
    subtitle = "Lighter cars generally travel farther per gallon",
    x = "Weight (1,000 lbs)",
    y = "Miles per gallon",
    fill = "Cylinders",
    shape = "Transmission",
    caption = "Source: mtcars dataset"
  ) +
  guides(
    fill = guide_legend(
      order = 1,
      override.aes = list(shape = 21, color = "#333333", size = 4, alpha = 1)
    ),
    shape = guide_legend(
      order = 2,
      override.aes = list(fill = "#BFC7C9", color = "#333333", size = 4, alpha = 1)
    )
  )
```

## 2. Cylinders and Fuel Economy Distribution

The second chart shifts from a numeric relationship to a group comparison. The fuel economy distribution shows that four-cylinder models reach much higher mpg values than eight-cylinder models.

```{r}
#| label: fig-mpg-by-cylinders
#| fig-width: 7.2
#| fig-height: 5
#| out-width: "100%"
#| fig-cap: "Fuel economy distribution by cylinder group. Points show the individual car models."

ggplot(cars_data, aes(x = cylinders, y = efficiency_mpg, fill = cylinders)) +
  geom_boxplot(
    width = 0.55,
    alpha = 0.9,
    outlier.shape = NA,
    color = "#333333",
    linewidth = 0.7,
    show.legend = FALSE
  ) +
  geom_jitter(
    aes(shape = transmission, fill = cylinders),
    width = 0.13,
    size = 3.4,
    stroke = 0.8,
    alpha = 0.95,
    color = "white"
  ) +
  scale_fill_manual(values = cyl_palette) +
  scale_shape_manual(values = c("Automatic" = 21, "Manual" = 24)) +
  scale_y_continuous(breaks = seq(10, 35, 5)) +
  coord_cartesian(ylim = c(9.5, 35.5)) +
  guides(
    fill = guide_legend(
      order = 1,
      override.aes = list(shape = 21, color = "#333333", size = 4, alpha = 1)
    ),
    shape = guide_legend(
      order = 2,
      override.aes = list(fill = "#BFC7C9", color = "#333333", size = 4, alpha = 1)
    )
  ) +
  labs(
    title = "Fuel Economy by Cylinder Count",
    subtitle = "Four-cylinder cars have the strongest mpg distribution",
    x = "Number of cylinders",
    y = "Miles per gallon",
    fill = "Cylinders",
    shape = "Transmission",
    caption = "Source: mtcars dataset"
  )
```

## 3. Average Horsepower by Group

The third chart explains the tradeoff: eight-cylinder cars have much higher average horsepower, but that choice comes with higher weight and lower fuel economy.

```{r}
#| label: fig-average-horsepower
#| fig-width: 7.2
#| fig-height: 4.8
#| out-width: "100%"
#| fig-cap: "Average horsepower by cylinder group. Horsepower is measured in hp."

ggplot(engine_summary, aes(x = cylinders, y = avg_hp, fill = cylinders)) +
  geom_col(width = 0.58, color = "white", linewidth = 1) +
  scale_fill_manual(values = cyl_palette) +
  scale_y_continuous(
    breaks = seq(0, 225, 50),
    expand = expansion(mult = c(0, 0.06))
  ) +
  coord_cartesian(ylim = c(0, 225)) +
  guides(
    fill = guide_legend(
      order = 1,
      override.aes = list(shape = 21, color = "#333333", size = 4, alpha = 1)
    )
  ) +
  labs(
    title = "Average Horsepower by Cylinder Count",
    subtitle = "More cylinders bring more power, with a visible efficiency tradeoff",
    x = "Number of cylinders",
    y = "Average horsepower",
    fill = "Cylinders",
    caption = "Source: mtcars dataset"
  )
```

## Conclusion and Call to Action

The three charts point to the same conclusion from different angles:

- vehicle weight is strongly associated with lower fuel economy;
- four-cylinder cars are generally lighter and more fuel-efficient;
- eight-cylinder cars offer more horsepower, but pay a clear mpg penalty.

**Call to action:** if the goal is to choose or design a fuel-efficient car, start with lighter models and fewer cylinders. Extra horsepower from an eight-cylinder engine should be justified by a real need, because in `mtcars` it comes with a clear fuel-economy cost.
