---
title: "Time Series"
format: html
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## 5.1 Discover datasets

In this exercise I explore some datasets available in R for time series
analysis. I classify them by periodicity, object class, and number of
observed variables.

### Base R: dataset list

```{r}
all_datasets <- data(package = "datasets")
all_datasets$results[, c("Package", "Item", "Title")]
```

```{r}
idx <- grepl(
  "air|temp|death|gas|stock|sun|rain|sales|passenger|co2|nile|market",
  all_datasets$results[, "Item"],
  ignore.case = TRUE
)
all_datasets$results[idx, c("Package", "Item", "Title")]
```

### Base R: datasets that look like time series

```{r}
period_from_frequency <- function(freq) {
  if (is.na(freq)) {
    "not a ts object"
  } else if (freq == 1) {
    "yearly"
  } else if (freq == 4) {
    "quarterly"
  } else if (freq == 12) {
    "monthly"
  } else if (freq == 260) {
    "something else: trading days"
  } else {
    paste0("something else: frequency ", freq)
  }
}

describe_dataset <- function(name, package = "datasets") {
  env <- new.env()
  data(list = name, package = package, envir = env)
  x <- get(name, envir = env)

  columns <- if (is.null(dim(x))) 1L else ncol(x)

  data.frame(
    dataset = name,
    object_class = paste(class(x), collapse = ", "),
    frequency = if (is.ts(x)) frequency(x) else NA,
    periodicity = if (is.ts(x)) period_from_frequency(frequency(x)) else "not a ts object",
    start = if (is.ts(x)) paste(start(x), collapse = " Q/M ") else NA,
    end = if (is.ts(x)) paste(end(x), collapse = " Q/M ") else NA,
    variables = columns,
    type = if (columns == 1L) "univariate" else "multivariate",
    row.names = NULL
  )
}

base_selected <- c(
  "AirPassengers",
  "JohnsonJohnson",
  "Nile",
  "EuStockMarkets",
  "co2"
)

do.call(rbind, lapply(base_selected, describe_dataset))
```

Summary:

| Dataset | Periodicity | Univariate or multivariate | What it measures |
|-----------------|--------------------:|-----------------|-----------------|
| `AirPassengers` | Monthly | Univariate | Monthly international airline passenger numbers, 1949-1960. |
| `JohnsonJohnson` | Quarterly | Univariate | Quarterly earnings per Johnson & Johnson share. |
| `Nile` | Yearly | Univariate | Annual flow of the River Nile. |
| `EuStockMarkets` | Other: trading days | Multivariate | Daily closing prices of four major European stock indices. |
| `co2` | Monthly | Univariate | Atmospheric CO2 concentration at Mauna Loa. |

At least three datasets that clearly contain time series data are
`AirPassengers`, `JohnsonJohnson`, and `Nile`. `EuStockMarkets` and
`co2` are also valid examples.

### Datasets selected from `timeSeriesDataSets`

The package is not required to read this table. If the package is
installed, the next chunk also checks the actual class with `class()`.

```{r}
time_series_datasets_selected <- data.frame(
  package = "timeSeriesDataSets",
  dataset = c("qgas_ts", "elecdaily_mts"),
  documented_class = c("ts", "mts / multivariate ts"),
  periodicity = c("quarterly", "something else: daily observations"),
  measures = c(
    "Australian gas production, in terajoules",
    "Electricity demand, workday indicator, and average temperature in Victoria, Australia"
  ),
  variables = c(1, 3),
  type = c("univariate", "multivariate")
)

time_series_datasets_selected
```

```{r}
if (requireNamespace("timeSeriesDataSets", quietly = TRUE)) {
  package_classes <- do.call(
    rbind,
    lapply(time_series_datasets_selected$dataset, describe_dataset,
           package = "timeSeriesDataSets")
  )
  package_classes
} else {
  "Package timeSeriesDataSets is not installed in this environment."
}
```

Interpretation:

| Dataset | Class | Periodicity | What it measures | Type |
|---------------|---------------|---------------|---------------|---------------|
| `qgas_ts` | `ts` | Quarterly | Australian gas production, measured in terajoules. | Univariate |
| `elecdaily_mts` | `mts`, a multivariate time series | Other: daily observations | Electricity demand, workday indicator, and daily average temperature in Victoria, Australia. | Multivariate |

### Datasets selected from `tsibbledata`

Again, the table is built so that it remains readable even if the
package is not installed. If `tsibbledata` is installed, the chunk
checks the class with `class()`.

```{r}
tsibbledata_selected <- data.frame(
  package = "tsibbledata",
  dataset = c("aus_production", "PBS"),
  documented_class = c("tsibble / tbl_ts", "tsibble / tbl_ts"),
  periodicity = c("quarterly", "monthly"),
  measures = c(
    "Selected Australian manufacturing production indicators: beer, tobacco, bricks, cement, electricity, gas",
    "Australian Medicare prescription scripts and script cost, disaggregated by concession, payment type, and ATC codes"
  ),
  variables = c(6, 2),
  type = c("multivariate", "multivariate")
)

tsibbledata_selected
```

```{r}
if (requireNamespace("tsibbledata", quietly = TRUE)) {
  package_classes <- do.call(
    rbind,
    lapply(tsibbledata_selected$dataset, describe_dataset,
           package = "tsibbledata")
  )
  package_classes
} else {
  "Package tsibbledata is not installed in this environment."
}
```

Interpretation:

| Dataset | Class | Periodicity | What it measures | Type |
|---------------|---------------|---------------|---------------|---------------|
| `aus_production` | `tsibble` / `tbl_ts` | Quarterly | Australian production of beer, tobacco, bricks, cement, electricity, and gas. | Multivariate |
| `PBS` | `tsibble` / `tbl_ts` | Monthly | Australian Medicare prescription counts and script costs. | Multivariate |

## 5.2 Recognize time series objects

```{r}
series_list <- list(
  AirPassengers = AirPassengers,
  JohnsonJohnson = JohnsonJohnson,
  nottem = nottem,
  ldeaths = ldeaths,
  EuStockMarkets = EuStockMarkets
)

lapply(series_list, class)
lapply(series_list, start)
lapply(series_list, end)
lapply(series_list, frequency)
```

```{r}
frequency_label <- function(freq) {
  if (freq == 1) {
    "yearly"
  } else if (freq == 4) {
    "quarterly"
  } else if (freq == 12) {
    "monthly"
  } else if (freq == 260) {
    "trading days"
  } else {
    paste("frequency", freq)
  }
}

series_summary <- data.frame(
  dataset = names(series_list),
  object_class = vapply(series_list, function(x) paste(class(x), collapse = ", "), character(1)),
  is_ts_object = vapply(series_list, inherits, logical(1), "ts"),
  start = vapply(series_list, function(x) paste(start(x), collapse = ", "), character(1)),
  end = vapply(series_list, function(x) paste(end(x), collapse = ", "), character(1)),
  frequency = vapply(series_list, frequency, numeric(1)),
  periodicity = vapply(series_list, function(x) frequency_label(frequency(x)), character(1)),
  variables = vapply(series_list, function(x) if (is.null(dim(x))) 1L else ncol(x), integer(1)),
  type = vapply(
    series_list,
    function(x) if (is.null(dim(x)) || ncol(x) == 1L) "univariate" else "multivariate",
    character(1)
  ),
  row.names = NULL
)

series_summary
```

Answer:

| Dataset | Stored as `ts` object? | Univariate or multivariate | Frequency |
|---------------|---------------|---------------|---------------|
| `AirPassengers` | Yes: class `ts` | Univariate | 12, monthly |
| `JohnsonJohnson` | Yes: class `ts` | Univariate | 4, quarterly |
| `nottem` | Yes: class `ts` | Univariate | 12, monthly |
| `ldeaths` | Yes: class `ts` | Univariate | 12, monthly |
| `EuStockMarkets` | Yes: classes `mts`, `ts`, `matrix`, `array` | Multivariate | 260, trading days |

## 5.3 Play with a time series dataset

I chose `AirPassengers`.

```{r}
x <- AirPassengers

class(x)
start(x)
end(x)
frequency(x)
summary(x)
plot(x, col = "steelblue", main = deparse(substitute(x)))
```

The series rises from 1949 to 1960, so it has a visible upward trend.
Passenger numbers follow the same peaks and dips each year, which shows
strong monthly seasonality. The seasonal swings widen as the level rises,
and the plot does not show one isolated outlier or sudden break.

## 5.4 Play with a dataset from `timeSeriesDataSets`

I chose `gtemp_both_ts`.

```{r}
if (requireNamespace("timeSeriesDataSets", quietly = TRUE)) {
  library(timeSeriesDataSets)

  x <- gtemp_both_ts

  class(x)
  start(x)
  end(x)
  frequency(x)
  summary(x)
  plot(x, col = "firebrick", main = deparse(substitute(x)))
} else {
  message("Install 'timeSeriesDataSets' to complete this exercise.")
}
```

`gtemp_both_ts` measures global mean land and open ocean temperature deviations
from the long-term average, in degrees Celsius. It is univariate because it
contains one numeric series. The plot should show annual variation around a
rising path, with the strongest increase near the end of the series. It
qualifies as a time series because the observations are ordered by year, stored
as a `ts` object, and sampled with frequency 1.

## 5.5 Play with a dataset from `tsibbledata`

I chose `gafa_stock`.

```{r}
# Uncomment if needed:
# install.packages(c("tsibbledata", "ggplot2"))

if (requireNamespace("tsibbledata", quietly = TRUE)) {
  library(tsibbledata)
  library(ggplot2)

  x <- gafa_stock

  class(x)
  head(x)

  ggplot(x, aes(x = Date, y = Close, colour = Symbol)) +
    geom_line() +
    labs(
      title = "Example tsibbledata dataset",
      x = "Date",
      y = "Close"
    )
} else {
  message("Install 'tsibbledata' to complete this exercise.")
}
```

The time index is `Date`, and `Symbol` separates Google, Amazon, Facebook, and
Apple. The measured variables are `Open`, `High`, `Low`, `Close`, `Adj_Close`,
and `Volume`; the plot uses `Close`. Amazon and Google trade at higher closing
prices than Apple and Facebook for much of the period, and the lines move with
sharp market swings rather than a seasonal pattern. This differs from a basic
`ts` object because the data keep dates, stock keys, and several measured
columns in one table, while a basic `ts` usually stores values in a regular
numeric vector or matrix with an implicit time scale.

## 5.6 Alternative representations, same same data

```{r}
values <- c(120, 135, 142, 160, 175)
values

values_ts <- ts(values, start = 2020, frequency = 1)
values_ts

class(values)
class(values_ts)

plot(values_ts,
     type = "o",
     pch = 16,
     col = "darkgreen",
     main = "Values as a Time Series",
     xlab = "Year",
     ylab = "Value")
```

The plain vector stores only the five numbers. `values_ts` stores the same
numbers plus the start year, yearly frequency, and `ts` class, so R can label
the observations as 2020 to 2024. The `ts` object is more useful for time-aware
analysis because plots and time-series functions can use years instead of raw
positions 1 to 5.

## 5.7 Why time matters

```{r}
set.seed(123)

x <- AirPassengers
x_shuffled <- ts(
  sample(as.numeric(x)),
  start = start(x),
  frequency = frequency(x)
)

par(mfrow = c(1, 2))
plot(x, col = "steelblue", main = "Original series")
plot(x_shuffled, col = "tomato", main = "Shuffled series")
par(mfrow = c(1, 1))
```

```{r}
par(mfrow = c(1, 2))
acf(x, main = "ACF: original")
acf(x_shuffled, main = "ACF: shuffled")
par(mfrow = c(1, 1))
```

Yes, the two series contain the same passenger counts. They look different
because shuffling assigns those counts to different months, so the upward trend
and yearly peaks no longer line up. The original ACF shows strong dependence
across nearby months and seasonal lags; the shuffled ACF drops close to random
noise because the order has been broken. This shows that temporal order carries
trend, seasonality, and autocorrelation, not just the values themselves.

## 5.8 A simple moving average

I chose `AirPassengers`.

```{r}
x <- AirPassengers

x_ma <- stats::filter(x, rep(1/12, 12), sides = 2)

plot(x,
     main = "Series with 12-period moving average",
     xlab = "Time",
     ylab = "Value",
     col = "gray40")
lines(x_ma, col = "red", lwd = 2)
```

The 12-month moving average makes the long-run rise in passenger numbers easier
to see. It is smoother than the original series because each point averages a
full year of monthly values. The month-to-month jumps and regular summer peaks
become less visible, while the underlying growth remains clear.

## 5.9 A forecasting experiment

I chose `AirPassengers`.

```{r}
# Uncomment if needed:
# install.packages("forecast")

library(forecast)

x <- AirPassengers

fc_naive <- naive(x, h = 12)
plot(fc_naive, main = "Naive forecast")
```

```{r}
fc_rw <- rwf(x, h = 12)
fc_snaive <- snaive(x, h = 12)

plot(fc_rw, main = "Random walk forecast")
plot(fc_snaive, main = "Seasonal naive forecast")
```

The naive forecast assumes every future value equals the last observed value.
For `AirPassengers`, that flat forecast is weak because the series has a strong
monthly pattern and a rising level. `rwf()` gives the same basic random-walk
idea here. `snaive()` fits this dataset best among the three because it repeats
last year's monthly pattern, which matches the visible seasonality.

## 5.10 Search and play with custom time series dataset

I chose `airline-passengers.csv` from `jbrownlee/Datasets`.

```{r}
url <- "https://raw.githubusercontent.com/jbrownlee/Datasets/master/airline-passengers.csv"
custom_df <- tryCatch(
  read.csv(url),
  error = function(e) {
    message("Could not read the GitHub URL; using an offline copy from AirPassengers.")
    data.frame(
      Month = format(
        seq(as.Date("1949-01-01"), by = "month", length.out = length(AirPassengers)),
        "%Y-%m"
      ),
      Passengers = as.numeric(AirPassengers)
    )
  }
)
head(custom_df)
str(custom_df)
```

```{r}
plot(custom_df[[2]],
     type = "l",
     col = "steelblue",
     main = "Custom dataset from jbrownlee/Datasets",
     xlab = "Observation index",
     ylab = names(custom_df)[2])
```

```{r}
custom_ts <- ts(custom_df[[2]], start = c(1949, 1), frequency = 12)
custom_ts

plot(custom_ts,
     col = "steelblue",
     main = "Airline passengers as a ts object",
     xlab = "Year",
     ylab = "Passengers")
```

The file measures monthly airline passenger counts. The first column stores
months, so frequency 12 makes sense for the `ts` conversion. The series rises
from 1949 to 1960 and repeats the same yearly peaks. The seasonal swings grow
larger as passenger numbers increase.

## 5.11 Student choice mini-investigation

I chose `JohnsonJohnson`.

```{r}
x <- JohnsonJohnson

class(x)
start(x)
end(x)
frequency(x)
summary(x)
plot(x, col = "steelblue", main = deparse(substitute(x)))
```

Dataset: `JohnsonJohnson`. It measures quarterly earnings per Johnson &
Johnson share from 1960 Q1 to 1980 Q4. The object class is `ts`, and frequency
4 means four observations per year. The series climbs over time, with small
values in the 1960s and much higher values by the late 1970s. It also shows a
repeated quarterly pattern. Variability grows as the level rises, so later
swings look larger than early swings. I do not see one isolated outlier, but
the last years move sharply. The series looks suitable for forecasting if the
model handles trend, quarterly seasonality, and changing variance.

## 5.12 Compare several time series from different sources

I chose `JohnsonJohnson` from base R, `gtemp_both_ts` from
`timeSeriesDataSets`, and the airline passenger CSV converted to `custom_ts`.

```{r}
if (requireNamespace("timeSeriesDataSets", quietly = TRUE)) {
  library(timeSeriesDataSets)
  data("gtemp_both_ts", package = "timeSeriesDataSets")
} else {
  message("Install 'timeSeriesDataSets' to use gtemp_both_ts.")
}

if (!exists("custom_ts")) {
  custom_ts <- ts(as.numeric(AirPassengers), start = c(1949, 1), frequency = 12)
}

par(mfrow = c(3, 1))
plot(JohnsonJohnson, col = "steelblue", main = "JohnsonJohnson")
if (exists("gtemp_both_ts")) {
  plot(gtemp_both_ts, col = "firebrick", main = "gtemp_both_ts")
}
plot(custom_ts, col = "darkorange", main = "airline-passengers.csv")
par(mfrow = c(1, 1))
```

| Series | Source | What it measures | Frequency | Most visible feature |
|---------------|---------------|---------------|---------------|---------------|
| `JohnsonJohnson` | base R | Quarterly earnings per Johnson & Johnson share | Quarterly | Upward trend with quarterly movement and larger late swings |
| `gtemp_both_ts` | `timeSeriesDataSets` | Global mean land and open ocean temperature deviations | Yearly | Long rise, strongest near the end |
| `custom_ts` | GitHub CSV | Monthly airline passenger counts | Monthly | Strong upward trend with repeated yearly peaks |

`gtemp_both_ts` shows the clearest long-run trend because the rise dominates the
plot. `custom_ts` shows the clearest seasonality, with peaks returning each
year. `JohnsonJohnson` looks hardest to forecast by eye because it combines
trend, quarterly movement, and widening swings.

## Final questions

### What distinguishes a time series from an ordinary unordered dataset?

A time series records observations in a meaningful time order. In an unordered
dataset, row order usually carries no signal. In a time series, row order is
part of the information.

### Why does temporal ordering matter?

Temporal ordering matters because each observation can depend on earlier
observations. The order reveals trend, seasonality, cycles, shocks, and
autocorrelation. Shuffling the same values can erase those patterns.

### What is the difference between a numeric vector, a data frame with a time column, and a `ts` object?

A numeric vector stores values only. A data frame with a time column stores
values and explicit dates, which makes it flexible for irregular data or
multiple variables. A `ts` object stores values with a start time and frequency,
so R can treat the data as a regular time series.

### What are some usual objectives of time series analysis?

Common goals include describing trend and seasonality, measuring dependence
between observations, forecasting future values, and spotting unusual changes.
Time series analysis can also compare series across groups or study how shocks
move through time.

### Why can two datasets contain the same values but still convey very different information when ordered differently?

The same values can tell a different story when their positions change. A rising
sequence suggests growth, while the same values in random order look unstable
and patternless. Order controls the visible trend, seasonal structure, and
autocorrelation.
